Computing volume using irregularly shaped rectangular prisms

### 2019-08-21 Cav2D/article-template.tex
title change
fixed small bugs

\title{Computing volume via irregularly shaped rectangular prisms}
%\author{Woodrow Wilson and Herbert Hoover} %%leave blank in initial submission to allow for double blind reviewing
%\author{ }

@@

\noindent
Two planar figures or solids are Cavalieri congruent if they can be decomposed into the same set of parallel ``indivisibles''. An indivisible of 
a planar figure is any cord of the planar figure. An indivisible of a solid is any planar section of a solid. Two indivisibles are equal if they are of the same length or area.
It is Cavalieri's principle that assures us that two planar figures or solids that are composed of the same set of indivisibles have equal area or volume \cite{andersen1985,eves1991,jullien2015,malik1984,wildberger2002,william1917,young1985}. Cavalieri's principle 
in $\mathbb{R}^3$ is stated below without proof:

@@

\noindent
Moreover, assume that $\mathbf{c}(t)$ is a translational curve with respect to $f(x)$ on $[a,b]$. Let us first partition $[a,b]$ into $n$ subintervals $[x_{i},x_{i+1}]$ of equal 
width $\Delta x^1 = \Delta x_i^1 = x_{i+1}-x_i\ll (b-a)$. Furthermore, let the transformation function $h$ map the $x$--coordinate $x_i^1\in [a,b]$ at which $\mathbf{c}(t) + \left <x_i^1,0 \right >$ intersects the 
$x$-axis to the $x$--coordinate at which $\mathbf{c}(t) + \left <x_i^1,0 \right >$ intersects $f(x)$. Note, the domain of $h$ is $[a,b]$. Let $a'$ denote the $x$-coordinate 
at which $\mathbf{c}(t) + \left <a,0 \right >$ intersects $f(x)$ and let $b'$ denote the $x$-coordinate 
at which $\mathbf{c}(t) + \left <b,0 \right >$ intersects $f(x)$.  Note, the range 
of $h$ is $[a',b']$. As the partition of $[a,b]$ can be made arbitrarily fine, $h:[a,b]\rightarrow [a',b']$ actually maps every point $x_i^1\in [a,b]$ to a point $x_i^2\in [a',b']$. \\

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{fig13.eps}
\caption{Region bounded by the $x$-axis and the lines $f(x)=x$, $a(y)=1-y$, and $b(y)=4-y$. The parametric representation of $a(y)$ is $\mathbf{c}(t) = \frac{1}{2} \left <-t,t \right >$. The figure also depicts the partition points $x_i^1$ and $x_i^2$ as used in the Cavalieri sum (see Equation~\eqref{eq:caval1}). Reproduced from Quaestiones Mathematicae (2012) 35: 265-296 with permission \copyright~ NISC (Pty) Ltd.}
\label{fig:caval2}
\end{figure}

@@

\noindent
The only thing that now remains is to find an expression for $h$. Note that, for every coordinate $(x_i^1,x_i^2) \in [a,b] \times [a',b']$ there exists a unique $t\in\mathbb{R}$ such that
\begin{equation}
 \label{eq:pat_rel_1d}
 \left <x_i^2,f(x_i^2) \right > = \left <x_i^1,0 \right> + \left<c_x(t),c_y(t) \right >.
 \end{equation}
 We can rewrite Equation~\eqref{eq:pat_rel_1d} as
 \begin{equation}
 \label{eq:t_func_1d}
 c_y(t) = f(x_i^1 + c_x(t)).
 \end{equation}
 We can now solve for $t$ in Equation~\eqref{eq:t_func_1d} in terms of $x_i^1$. If we substitute the solution $t(x_i)$ so obtained into Equation~\eqref{eq:pat_rel_1d} we obtain 
\begin{equation}
\label{eq:h_func_1d}
 x_i^2 = h(x_i^1) = x_i^1 + c_x\circ t(x_i^1).
\end{equation}

@@

\noindent
Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$.  Let us further assume that $f(x,y) > 0$ for all $(x,y)\in S$. Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and 
$S_f$ via translations of $\mathbf{c}(t)$ along the $xy$ plane and then closing the resulting open faces that form due to the aforementioned connections by employing appropriate surfaces. Note that 
$\mathcal{S}$ can be decomposed into an infinite number of irregularly shaped rectangular infinitesimals. As an example, the solid $\mathcal{S}$ depicted in Figure~\ref{fig:prismatoid_solid} is obtained if we set $f(x,y) = -x - y + 8$, $\mathbf{c}(t) = \left<\frac{1}{2}t,t,t\right>$ and $R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber$.
Moreover, the domain $R$ and the range $S$ of the transformation function $T$ which is associated with the solid $\mathcal{S}$ is depicted in Figure~\ref{fig:prismatoid_regions}.  
In general, let $V_{\mathcal{S}}$ denote the volume of $\mathcal{S}$. We can approximate $V_{\mathcal{S}}$ with:
\begin{equation}
\label{eq:cav_sum_2d}
V_{\mathcal{S}} \approx \sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta x_i^1 \Delta y_j^1.
\end{equation}
Let us verify that this is indeed the case. If we closely inspect Equation~\eqref{eq:cav_sum_2d} and Figure~\ref{fig:prismatoid_solid} we notice the following:
\begin{itemize}
\item $\Delta x_i^1$ and $\Delta y_j^1$ represent the width and the length of subrectangle $R_{ij}$.
\item The product $\Delta A = \Delta x_i^1\times \Delta y_j^1$ represents the area of subrectangle $R_{ij}$.
\item $f(x_i^2,y_j^2)$ represents the height of an irregular rectangular column or prism $\mathcal{K}_{ij}$ (or in the case of Figure~\ref{fig:prismatoid_solid} and oblique rectangular prism) whose base is $R_{ij}$. Furthermore, the lateral edges of this column are all translations 
of $\mathbf{c}(t)$ along the $xy$-plane. The cross-section between any parallel plane to the $xy$-plane and $\mathcal{K}_{ij}$ is a rectangle with the same dimensions as $R_{ij}$. 
\item The product $f(x_i^2,y_j^2)\times \Delta A$ represents the volume of $\mathcal{K}_{ij}$ (due to Cavalieri's principle).
\item The double Cavalieri sum $\sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta A$, therefore, approximates $V_{\mathcal{S}}$. 
\item In the limit the double Cavalieri sum approaches $V_{\mathcal{S}}$ (if this limit exists). 
\end{itemize}

@@

\noindent
Since $\mathcal{S}$ can be decomposed into an infinite number of irregularly shaped rectangular infinitesimals we can compute its volume using:
\begin{equation}
\label{eq:cav_ex2}
V_{\mathcal{S}} = \iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We, however, first need to calculate $T$ if we wish to evaluate Equation~\eqref{eq:cav_ex2} (see Figure~\ref{fig:sphere_regions}). We can determine $T$ via 
Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d}. Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} in turn requires $t(x_i^1,y_j^2)$, which can be obtained via Equation~\eqref{eq:c_z}.
If we substite $f(x,y)$ and $\mathbf{c}(t)$ into Equation~\eqref{eq:c_z} and then solve for $t$ we obtain:
\begin{equation}
t(x_{i}^1,y_{j}^1) = \frac{1}{4}[-2 x_{i}^1 - 2 y_{j}^1  -  1 + K(x_{i}^1,y_{j}^1)],\nonumber
\end{equation}
where
\begin{equation}
K(x_{i}^1,y_{j}^1) = \sqrt{-4(x_{i}^1)^2+4x_{i}^1+8x_{i}^1y_{j}^1-4(y_{j}^1)^2+4y_{j}^1 + 33}.\nonumber
\end{equation}
As mentioned earlier, we can use $t(x_i^1,y_j^1)$ to find expressions for $h$ and $s$. The expressions of $h$ and $s$ so obtained can then be used to evaluate Equation~\eqref{eq:cav_ex2}:  
\begin{eqnarray}
\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \int_0^1\int_0^1 f(h(x,y),s(x,y))~dydx\nonumber\\
&=& \int_0^1 \int_0^1 \sqrt{-(x+t(x,y))^2 - (y+t(x,y))^2 + 4} ~dydx\nonumber\\
&=& 0.864043.\nonumber
\end{eqnarray}

>>>>

\title{Computing volume using irregularly shaped rectangular prisms}
%\author{Woodrow Wilson and Herbert Hoover} %%leave blank in initial submission to allow for double blind reviewing
%\author{ }

@@

\noindent
Two planar figures or solids are Cavalieri congruent if they can be decomposed into the same set of parallel ``indivisibles''. An indivisible of 
a planar figure is any cord of the planar figure. An indivisible of a solid is any planar section of a solid. Two indivisibles are equal if they are of the same length or area.
Cavalieri's principle assures us that two planar figures or solids that are composed of the same set of indivisibles have equal area or volume \cite{andersen1985,eves1991,jullien2015,malik1984,wildberger2002,william1917,young1985}. Cavalieri's principle 
in $\mathbb{R}^3$ is stated below without proof:

@@

\noindent
Moreover, assume that $\mathbf{c}(t)$ is a translational curve with respect to $f(x)$ on $[a,b]$. Let us first partition $[a,b]$ into $n$ subintervals $[x_{i},x_{i+1}]$ of equal 
width $\Delta x^1 = \Delta x_i^1 = x_{i+1}-x_i\ll (b-a)$. Furthermore, let the transformation function $h$ map the $x$--coordinate $x_i^1\in [a,b]$ at which $\mathbf{c}(t) + \left <x_i^1,0 \right >$ intersects the 
$x$-axis to the $x$--coordinate at which $\mathbf{c}(t) + \left <x_i^1,0 \right >$ intersects $f(x)$. Note, the domain of $h$ is $[a,b]$. Let $a'$ denote the $x$-coordinate 
at which $\mathbf{c}(t) + \left <a,0 \right >$ intersects $f(x)$ and let $b'$ denote the $x$-coordinate 
at which $\mathbf{c}(t) + \left <b,0 \right >$ intersects $f(x)$.  Note, the range 
of $h$ is $[a',b']$ (assuming $h$ is monotone increaing). As the partition of $[a,b]$ can be made arbitrarily fine, $h:[a,b]\rightarrow [a',b']$ actually maps every point $x_i^1\in [a,b]$ to a point $x_i^2\in [a',b']$. \\

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{fig13.eps}
\caption{Region bounded by the $x$-axis and the lines $f(x)=x$, $a(y)=1-y$, and $b(y)=4-y$. The parametric representation of $a(y)$ is $\mathbf{c}(t) = \frac{1}{2} \left <-t,t \right >+\left<a,0\right>$. The figure also depicts the partition points $x_i^1$ and $x_i^2$ as used in the Cavalieri sum (see Equation~\eqref{eq:caval1}). Reproduced from Quaestiones Mathematicae (2012) 35: 265-296 with permission \copyright~ NISC (Pty) Ltd.}
\label{fig:caval2}
\end{figure}

@@

\noindent
The only thing that now remains is to find an expression for $h$. Note that, for every coordinate $(x_i^1,x_i^2) \in [a,b] \times [a',b']$ there exists a unique $t\in\mathbb{R}$ such that
\begin{equation}
 \label{eq:pat_rel_1d}
 \left <x_i^2,f(x_i^2) \right > = \left <x_i^1,0 \right> + \left<c_x(t),c_y(t) \right >.
 \end{equation}
 We can rewrite Equation~\eqref{eq:pat_rel_1d} as
 \begin{equation}
 \label{eq:t_func_1d}
 c_y(t) = f(x_i^1 + c_x(t)).
 \end{equation}
 We can now solve for $t$ in Equation~\eqref{eq:t_func_1d} in terms of $x_i^1$. If we substitute the solution $t(x_i^1)$ so obtained into Equation~\eqref{eq:pat_rel_1d} we obtain 
\begin{equation}
\label{eq:h_func_1d}
 x_i^2 = h(x_i^1) = x_i^1 + c_x\circ t(x_i^1).
\end{equation}

@@

\noindent
Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$.  Let us further assume that $f(x,y) > 0$ for all $(x,y)\in S$. Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and 
$S_f$ via translations of $\mathbf{c}(t)$ along the $xy$ plane and then closing the resulting open faces that form due to the aforementioned connections by employing appropriate surfaces. Note that 
$\mathcal{S}$ so formed can per definition be decomposed into an infinite number of irregularly shaped rectangular infinitesimals. As an example, the solid $\mathcal{S}$ depicted in Figure~\ref{fig:prismatoid_solid} is obtained if we set $f(x,y) = -x - y + 8$, $\mathbf{c}(t) = \left<\frac{1}{2}t,t,t\right>$ and $R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber$.
Moreover, the domain $R$ and the range $S$ of the transformation function $T$ which is associated with the solid $\mathcal{S}$ is depicted in Figure~\ref{fig:prismatoid_regions}.  
In general, let $V_{\mathcal{S}}$ denote the volume of $\mathcal{S}$. We can approximate $V_{\mathcal{S}}$ with:
\begin{equation}
\label{eq:cav_sum_2d}
V_{\mathcal{S}} \approx \sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta x_i^1 \Delta y_j^1.
\end{equation}
Let us verify that this is indeed the case. If we closely inspect Equation~\eqref{eq:cav_sum_2d} and Figure~\ref{fig:prismatoid_solid} we notice the following:
\begin{itemize}
\item $\Delta x_i^1$ and $\Delta y_j^1$ represent the width and the length of subrectangle $R_{ij}$.
\item The product $\Delta A = \Delta x_i^1\times \Delta y_j^1$ represents the area of subrectangle $R_{ij}$.
\item $f(x_i^2,y_j^2)$ represents the height of an irregular rectangular column or prism $\mathcal{K}_{ij}$ (or in the case of Figure~\ref{fig:prismatoid_solid} and oblique rectangular prism) whose base is $R_{ij}$. Furthermore, the lateral edges of this column are all translations 
of $\mathbf{c}(t)$ along the $xy$-plane. The cross-section between any parallel plane to the $xy$-plane and $\mathcal{K}_{ij}$ is a rectangle with the same dimensions as $R_{ij}$. 
\item The product $f(x_i^2,y_j^2)\times \Delta A$ represents the volume of $\mathcal{K}_{ij}$ (due to Cavalieri's principle).
\item The double Cavalieri sum $\sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta A$, therefore, approximates $V_{\mathcal{S}}$. 
\item In the limit the double Cavalieri sum approaches $V_{\mathcal{S}}$ (if this limit exists). 
\end{itemize}

@@

\noindent
Since $\mathcal{S}$ can be decomposed into an infinite number of irregularly shaped rectangular infinitesimals we can compute its volume using:
\begin{equation}
\label{eq:cav_ex2}
V_{\mathcal{S}} = \iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We, however, first need to calculate $T$ if we wish to evaluate Equation~\eqref{eq:cav_ex2} (see Figure~\ref{fig:sphere_regions}). We can determine $T$ via 
Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d}. Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} in turn requires $t(x_i^1,y_j^1)$, which can be obtained via Equation~\eqref{eq:c_z}.
If we substite $f(x,y)$ and $\mathbf{c}(t)$ into Equation~\eqref{eq:c_z} and then solve for $t$ we obtain:
\begin{equation}
t(x_{i}^1,y_{j}^1) = \frac{1}{4}[-2 x_{i}^1 - 2 y_{j}^1  -  1 + K(x_{i}^1,y_{j}^1)],\nonumber
\end{equation}
where
\begin{equation}
K(x_{i}^1,y_{j}^1) = \sqrt{-4(x_{i}^1)^2+4x_{i}^1+8x_{i}^1y_{j}^1-4(y_{j}^1)^2+4y_{j}^1 + 33}.\nonumber
\end{equation}
As mentioned earlier, we can use $t(x_i^1,y_j^1)$ to find expressions for $h$ and $s$. The expressions of $h$ and $s$ so obtained can then be used to evaluate Equation~\eqref{eq:cav_ex2}:  
\begin{eqnarray}
\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \int_0^1\int_0^1 f(h(x,y),s(x,y))~dydx\nonumber\\
&=& \int_0^1 \int_0^1 \sqrt{-(x+t(x,y))^2 - (y+t(x,y))^2 + 4} ~dydx\nonumber\\
&=& 0.864043.\nonumber
\end{eqnarray}
